---
title: "Longitudinal Analysis of the ARMD Trial"
author: "Miguel Castro"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")
library(nlmeU)
library(nlme)
library(lattice)
```

## Data

Build a random 150-patient subsample of the ARMD trial and keep the long (post-baseline)
format.

```{r data}
set.seed(43)
idx <- sample(1:240, 150)
data(armd.wide)
data(armd)
sel <- levels(armd.wide$subject)[idx]
armd.wide.s <- droplevels(armd.wide[armd.wide$subject %in% sel, ])
armd.s      <- droplevels(armd[armd$subject %in% sel, ])  # long format, post-baseline only
```

Quick description and the missing-data pattern (dropout is mostly monotone and at the end of
follow-up; treated as MAR, models fitted by ML).

```{r describe}
dim(armd.wide.s)
table(armd.wide.s$treat.f)
table(armd.wide.s$miss.pat)
```

Mean and sd of visual acuity by visit and arm — note the sd grows over time (non-constant
variance).

```{r means}
with(armd.s, round(tapply(visual, list(time.f, treat.f), mean, na.rm = TRUE), 2))
with(armd.s, round(tapply(visual, list(time.f, treat.f), sd,   na.rm = TRUE), 2))
```

## Marginal models (gls)

Same mean structure as the ARMD case study: baseline as a covariate, a mean per visit, and a
treatment-by-visit interaction. Fit homoscedastic, then add a power-of-time variance and
within-subject correlation structures, comparing by likelihood-ratio tests.

```{r gls}
lm1.form <- formula(visual ~ -1 + visual0 + time.f + time.f:treat.f)

fm6.1 <- gls(lm1.form, data = armd.s)                                    # homoscedastic
fm9.2 <- gls(lm1.form, weights = varPower(form = ~time), data = armd.s)  # power-of-time variance
anova(fm6.1, fm9.2)

# add within-subject correlation: compound symmetry, then general/unstructured
fm12.1 <- gls(lm1.form, weights = varPower(form = ~time),
              correlation = corCompSymm(form = ~tp | subject), data = armd.s)
fm12.3 <- gls(lm1.form, weights = varPower(form = ~time),
              correlation = corSymm(form = ~tp | subject), data = armd.s)
anova(fm9.2, fm12.1, fm12.3)
intervals(fm12.3, which = "var-cov")$corStruct   # closer visits correlate more
```

## Mixed models (lme)

Attribute the within-subject correlation to subject-level random effects: random intercept,
then intercept + slope, with a power variance on the residuals.

```{r lme}
fm16.1 <- lme(lm1.form, random = ~1 | subject, data = armd.s)   # random intercept (ICC ~ 0.50)
VarCorr(fm16.1)

fm16.2 <- lme(lm1.form, random = ~1 | subject,
              weights = varPower(form = ~time), data = armd.s)   # + power variance
anova(fm16.1, fm16.2)

fm16.3 <- lme(lm1.form, random = ~1 + time | subject,           # + random slope (final model)
              weights = varPower(form = ~time), data = armd.s,
              control = lmeControl(opt = "optim", maxIter = 200, msMaxIter = 200))
anova(fm16.2, fm16.3)
summary(fm16.3)
```

## Treatment effect

The treatment effect is spread over four interaction terms, so test it jointly: refit by ML
and compare against a model with no treatment terms via a likelihood-ratio test.

```{r treatment}
fm.full <- update(fm16.3, method = "ML")
fm.red  <- update(fm.full, fixed = visual ~ -1 + visual0 + time.f)
anova(fm.red, fm.full)   # LR = 4.33, df = 4, p = 0.36 -> no significant treatment effect
```

## Diagnostics

```{r diagnostics}
plot(fm16.3, resid(., type = "pearson") ~ fitted(.) | time.f, abline = 0)
qqnorm(fm16.3, ~resid(., type = "pearson"))
qqnorm(fm16.3, ~ranef(.))
```
